Do Alternate Wetting and Drying Irrigation Technology and Nitrogen Rates Affect Rice Sheath Blight?

### 2018-11-02 analysis/paper/paper.Rmd
Correct formatting of superscript and dashes throughout

---
title: "Do Alternate Wetting and Drying Irrigation Technology and Nitrogen Rates Affect Rice Sheath Blight?"
author: |
  | Adam H. Sparks^[Corresponding author: Adam.Sparks@usq.edu.au]  \textsuperscript{1\dag}, Nancy P. Castilla \textsuperscript{1}, Bjorn Ole Sander \textsuperscript{1}
  | \textsuperscript{1}International Rice Research Institute, Crop and Environmental Sciences Division, Los Baños, Laguna, Philippines, 4031
  | \textsuperscript{\dag} Current Address: University of Southern Queensland, Centre for Crop Health, West Street, Tooowoomba, Queensland, Australia, 4350
csl: "../templates/european-journal-of-plant-pathology.csl" # Insert path for the bib-style
abstract: |
  Water and nitrogen management play vital roles in rice production. However, the mismanagement of these two management practices may trigger sheath blight of rice, caused by _Rhizoctonia solani_, which is favored by wet conditions, high relative humidity, and high nitrogen fertilizer levels. To understand how different combinations of water and nitrogen management affect sheath blight epidemics, we conducted two separate split-plot experiments with a water saving (alternate wetting and drying) regime and a traditional puddled regime combined with differing nitrogen treatments in the dry seasons of 2015 and 2016. Disease was scored in the same way in both experiments using a sheath blight assessment scale for field evaluation developed at the International Rice Research Institute to assess the severity on infected sheaths and leaves while sheath blight incidence on tillers were counted per hill. We were unable to detect any difference in the incidence of tiller sheath blight due to irrigation, tiller and leaf sheath blight did clearly differ statistically by irrigation treatment, but leaf sheath blight severity did not. Our findings suggest that farmers can adopt water saving technologies without risking increased sheath blight incidence. We suggest that further cross-cutting research in this area is warranted. All data generated from this project and code used for analysis are available as a reproducible research compendium from <https://github.com/openplantpathology/rice_awd_pests> [@Sparks2018].
bibliography: references.bib
output:
  bookdown::word_document2:
    toc: no
    reference_docx: "../templates/template.docx"
---

@@

Sheath blight disease (*Rhizoctonia solani* Kühn), anastomosis group 1 [(teleomorph: *Thanatephorus cucumeris* (A.B. Frank) Donk.] of rice is an economically important disease worldwide throughout tropical and temperate production areas. Under conducive conditions the disease can cause up to 50&nbsp;\% yield losses [@Marchetti1991]. Studies by Savary et al. [-@Savary2000] estimated normal yield losses in tropical irrigated lowland rice in Asia at 5-10 &nbsp;\%. The disease is best managed through an integrated disease management approach which includes resistance [@Kumar2009; @Srinivasachary2011] fungicide applications where affordable [@Rush1983; @Groth2008], nitrogen (N) management [@Slaton2003; @Castilla1996; @Tang2007] and water management [@Castilla1996].

Alternate wetting and drying (AWD) is an irrigation technique for rice (*Oryza sativa* L.) developed by the International Rice Research Institute (IRRI) and its partners that saves about 15-40&nbsp;\% of irrigation water while not affecting yield [@Bouman2001; @Feng2007]. In AWD rice, fields are exposed to several dry phases during the growing season without exposing the plants to water stress. In order to avoid yield decline under AWD “safe” thresholds have been developed. Under safe AWD irrigation water is applied when the field water level reaches 15&nbsp;cm below the soil surface [@Richards2014]. Fields are furthermore kept puddled during the flowering period to avoid plant stress. As an added benefit to saving water, AWD also reduces greenhouse gas (GHG) emissions of rice fields, which are a substantial factor in the GHG budget of rice producing countries, by around 50 &nbsp;\% [@Yan2005; @Sander2015].

The AWD technology has been identified as promising climate-smart practice for different rice growing regions that can stabilize rice production in water scarce areas as well as help reduce the carbon footprint of rice production. Various countries, *e.g.*, Bangladesh, Vietnam, Thailand and Philippines, plan to widely apply AWD to rice production [@MOEF2015]. However, a change like this in the water regime in rice fields on a large scale might trigger different other effects, for example changes in disease intensity or severity.

@@

The plot design was a split-plot randomised complete block design with four replicates where irrigation was the main plot treatment and N rate was the split-plot treatment.

The main plot treatments were alternate wetting and drying (AWD) and continuously puddled or farmers' practice as the control treatment. The main plot size was 12&nbsp;m x 12&nbsp;m (144&nbsp;m^2), with a split-plot size of 5&nbsp;m&nbsp;x&nbsp;5&nbsp;m (25&nbsp;m^2). Replication size was 12&nbsp;m&nbsp;x&nbsp;24&nbsp;m (288&nbsp;m^2) with a buffer of 1&nbsp;m per split-plot for a whole experiment size of 1,152&nbsp;m^2.

Irrigation in AWD plots was determined by the water level in plots, *i.e.*, when the water level reached 15&nbsp;cm below the soil surface irrigation water was applied to a level of 5&nbsp;cm. In puddled plots a standing water layer of 3-5&nbsp;cm was maintained throughout the growing season.

The split-plot treatment rates of N were N0 (no N supply), N100 (100&nbsp;kg&nbsp;per&nbsp;ha applied as urea in three splits) and N120 (120&nbsp;kg&nbsp;per&nbsp;ha applied as urea in three splits) (Tab.&nbsp;\@ref(tab:N_rate)).

@@

An isolate of *Rhizoctonia solani* AG1-1a previously isolated from infected rice was maintained on potato dextrose agar (PDA) medium in tubes. The isolate was transferred to 90&nbsp;mm Petri dishes containing PDA and incubated a room temperature (20 to 27&deg;C). Glass bottles of autoclaved rice grain and hull substrate were prepared and plugs of the culture were transferred from Petri dishes to the autoclaved substrate and incubated at room temperature for two weeks.

The plots were inoculated 20 days after transplanting by spreading 151&nbsp;g of inoculum over the split-plot area with a 1&nbsp;m buffer on either end (4&nbsp;m x 11&nbsp;m).

@@

The 2016 experiment followed the same split-plot design as described for the 2015 experiment. However, the plot size was increased and due to these changes, the replicate sizes are are not equal as necessitated by the use of a larger area for the experiment. The main plot sizes were: Block 1 (B1) 21&nbsp;m x 20.5&nbsp;m (412.5 sq&nbsp;m) and Block 2 (B2) 20.25&nbsp;m x 21.6&nbsp;m (437.4&nbsp;m^2). The split-plot sizes were B1 21&nbsp;m x 10.25&nbsp;m (215.25&nbsp;m^2), B2 20.25&nbsp;m x 10.8&nbsp;m (218.7&nbsp;m^2). The replication sizes were B1 - 42&nbsp;m&nbsp;x&nbsp;20.5&nbsp;m (861&nbsp;m^2) and B2 - 40.5&nbsp;m&nbsp;x&nbsp;21.6&nbsp;m (874.8&nbsp;m^2). A buffer 0.5&nbsp;m per split-plot was used and the overall experiment size was 3471.6&nbsp;m^2.

Split-plot N rates differed from the 2015 rates with only two N treatments, N60 (60&nbsp;kg&nbsp;/&nbsp;ha as urea split into two applications) and N180 (180&nbsp;kg&nbsp;/&nbsp;ha in three splits) being applied (Tab.&nbsp;\@ref(tab:N_rate)).

@@

In both experiments, two sample areas, sized 1&nbsp;m&nbsp;x&nbsp;1&nbsp;m, per plot were assessed. The total number of tillers per hill and number of tillers with sheath blight (incidence) were measured for nine hills per sample area. Tiller sheath severity was measured for four tillers per hill and six leaves tiller using a disease assessment rating scale for field evaluation developed at IRRI. The scale is an unequally spaced categorical scale where 0 - no disease; 1 - trace to 1&nbsp;\% severity; 2 - 1 to 5&nbsp;\% severity; 3 - 5 to 15&nbsp;\% severity; 4 - 15 to 50&nbsp;\% severity; 5 - 50 to 100&nbsp;\% severity. Severity was considered to be the amount of leaf or sheath tissue covered by sheath blight lesions. Leaves were rated as green (living) having at least XXX&nbsp;\% living tissue or dead, defined as having less than XXX&nbsp;\% living tissue. The same rating methodology was used in both experiments. Five disease assessments were made in the 2015 experiment and four disease assessments were made in the 2016 experiment, respectively.

@@

As most of the data's residuals did not meet assumptions for normality, the analysis was carried out using Bayesian multivariate generalised linear mixed models implemented in the R package `MCMCglmm` [@MCMCglmm2010]. Six models were created where the the dependent variables were 2015 tiller sheath blight incidence, 2015 tiller sheath blight severity, 2015 leaf sheath blight severity; 2016 tiller sheath blight incidence, 2016 tiller sheath blight severity; 2016 leaf sheath blight severity, respectively. The independent variables were the irrigation management and N rate with replicate treated as a random variable. The base levels (control treatments) for the analyses for 2015 were: N rate - N0, irrigation management - puddled and for 2016 they were: N rate - N60, irrigation management - puddled

Four MCMC chains were run for 55,000 steps with the first 5000 discarded as burn in. The remaining 50,000 samples were used to determine the posterior distribution of the parameters of the model. The `coda` [@coda2006] package was used to provide utilities to check MCMC chain convergence using visual inspection of resulting trace graphs and the Gelman-Rubin test.

@@

The combination of the irrigation and N rates did not appear to increase the disease severity or incidence in either experiment. In 2015 there were no clear differences in the interaction of irrigation and N rate between any of the three treatments. However, in the 2016 experiment the combination of AWD-N180 was clearly different, being lower, than the puddled-N60 combination treatment.

@@

Climate change is predicted to decrease sheath blight risk in the Korean peninsula over the longer term. However, over the next 10 - 20 years the risk is predicted to remain the same as from 2000 - 2010 [@kim2015predicting]. Therefore, it is likely to remain an important diease in the near future with little resistance available in current varieties. However, based on these studies the use of AWD technologies would not appear to be a factor that increases the incidence or severity of rice sheath blight disease.

@@

```{r fig-3, fig.cap="Sheath blight progress was measured in the 2015 at five assessment points and 2016 experiment at four points. Sheath blight incidence was rated as the number of infected tillers divided by the total number of tillers per hill and used to calculate the area under the disease progress stairs (AUDPS). Sheath blight severity was rated on an unevenly spaced categorical scale and converted to the mid-point percentage value to calculate the AUDPS. Main plot irrigation treatments were permanently puddled (PDL) and alternate wetting and drying (AWD)."}

@@

##### Tiller sheath blight incidence boxplot for N rate
```{r fig-4, fig.cap="Sheath blight severity progress was measured in the 2015 at five assessment points and 2016 experiment at four points. Sheath blight incidence was rated as the number of infected tillers divided by the total number of tillers per hill and used to calculate the area under the disease progress stairs (AUDPS). Sheath blight severity was rated on an unevenly spaced categorical scale and converted to the mid-point percentage value to calculate the AUDPS. Five different N treatments were applied to split plots depending on the experiment. In the 2015 experiment three N rate treatments were applied: no N (N0), 100&nbsp;kgkg&nbsp;/^&nbsp;ha N (N100) and 180&nbsp;kgkg&nbsp;/^&nbsp;ha N (N120). In the 2016 experiment, two N rate treatments were applied: 60&nbsp;kgkg&nbsp;/^&nbsp;ha N (N60) and 180&nbsp;kgkg&nbsp;/^&nbsp;ha N (N180)."}

>>>>

---
title: "Do Alternate Wetting and Drying Irrigation Technology and Nitrogen Rates Affect Rice Sheath Blight?"
author: |
  | Adam H. Sparks^[Corresponding author: Adam.Sparks@usq.edu.au]  \textsuperscript{1\dag}, Nancy P. Castilla \textsuperscript{1}, Bjorn Ole Sander \textsuperscript{1}
  | \textsuperscript{1}International Rice Research Institute, Crop and Environmental Sciences Division, Los Baños, Laguna, Philippines, 4031
  | \textsuperscript{\dag} Current Address: University of Southern Queensland, Centre for Crop Health, West Street, Tooowoomba, Queensland, Australia, 4350
csl: "../templates/european-journal-of-plant-pathology.csl" # Insert path for the bib-style
abstract: |
  Water and nitrogen management play vital roles in rice production. However, the mismanagement of these two management practices may trigger sheath blight of rice, caused by _Rhizoctonia solani_, which is favored by wet conditions, high relative humidity, and high nitrogen fertilizer levels. To understand how different combinations of water and nitrogen management affect sheath blight epidemics, we conducted two separate split--plot experiments with a water saving (alternate wetting and drying) regime and a traditional puddled regime combined with differing nitrogen treatments in the dry seasons of 2015 and 2016. Disease was scored in the same way in both experiments using a sheath blight assessment scale for field evaluation developed at the International Rice Research Institute to assess the severity on infected sheaths and leaves while sheath blight incidence on tillers were counted per hill. We were unable to detect any difference in the incidence of tiller sheath blight due to irrigation, tiller and leaf sheath blight did clearly differ statistically by irrigation treatment, but leaf sheath blight severity did not. Our findings suggest that farmers can adopt water saving technologies without risking increased sheath blight incidence. We suggest that further cross--disciplinary research in this area is warranted. All data generated from this project and code used for analysis are available as a reproducible research compendium from <https://github.com/openplantpathology/rice_awd_pests> [@Sparks2018].
bibliography: references.bib
output:
  bookdown::word_document2:
    toc: no
    reference_docx: "../templates/template.docx"
---

@@

Sheath blight disease (*Rhizoctonia solani* Kühn), anastomosis group 1 [(teleomorph: *Thanatephorus cucumeris* (A.B. Frank) Donk.] of rice is an economically important disease worldwide throughout tropical and temperate production areas. Under conducive conditions the disease can cause up to 50&nbsp;\% yield losses [@Marchetti1991]. Studies by Savary et al. [-@Savary2000] estimated normal yield losses in tropical irrigated lowland rice in Asia at 5--10 &nbsp;\%. The disease is best managed through an integrated disease management approach which includes resistance [@Kumar2009; @Srinivasachary2011] fungicide applications where affordable [@Rush1983; @Groth2008], nitrogen (N) management [@Slaton2003; @Castilla1996; @Tang2007] and water management [@Castilla1996].

Alternate wetting and drying (AWD) is an irrigation technique for rice (*Oryza sativa* L.) developed by the International Rice Research Institute (IRRI) and its partners that saves about 15--40&nbsp;\% of irrigation water while not affecting yield [@Bouman2001; @Feng2007]. In AWD rice, fields are exposed to several dry phases during the growing season without exposing the plants to water stress. In order to avoid yield decline under AWD “safe” thresholds have been developed. Under safe AWD irrigation water is applied when the field water level reaches 15&nbsp;cm below the soil surface [@Richards2014]. Fields are furthermore kept puddled during the flowering period to avoid plant stress. As an added benefit to saving water, AWD also reduces greenhouse gas (GHG) emissions of rice fields, which are a substantial factor in the GHG budget of rice producing countries, by around 50 &nbsp;\% [@Yan2005; @Sander2015].

The AWD technology has been identified as promising climate smart practice for different rice growing regions that can stabilize rice production in water scarce areas as well as help reduce the carbon footprint of rice production. Various countries, *e.g.*, Bangladesh, Vietnam, Thailand and Philippines, plan to widely apply AWD to rice production [@MOEF2015]. However, a change like this in the water regime in rice fields on a large scale might trigger different other effects, for example changes in disease intensity or severity.

@@

The plot design was a split--plot randomised complete block design with four replicates where irrigation was the main plot treatment and N rate was the split--plot treatment.

The main plot treatments were alternate wetting and drying (AWD) and continuously puddled or farmers' practice as the control treatment. The main plot size was 12&nbsp;m x 12&nbsp;m (144&nbsp;m^2^), with a split--plot size of 5&nbsp;m&nbsp;x&nbsp;5&nbsp;m (25&nbsp;m^2^). Replication size was 12&nbsp;m&nbsp;x&nbsp;24&nbsp;m (288&nbsp;m^2^) with a buffer of 1&nbsp;m per split--plot for a whole experiment size of 1,152&nbsp;m^2^.

Irrigation in AWD plots was determined by the water level in plots, *i.e.*, when the water level reached 15&nbsp;cm below the soil surface irrigation water was applied to a level of 5&nbsp;cm. In puddled plots a standing water layer of 3--5&nbsp;cm was maintained throughout the growing season.

The split--plot treatment rates of N were N0 (no N supply), N100 (100&nbsp;kg&nbsp;per&nbsp;ha applied as urea in three splits) and N120 (120&nbsp;kg&nbsp;per&nbsp;ha applied as urea in three splits) (Tab.&nbsp;\@ref(tab:N_rate)).

@@

An isolate of *Rhizoctonia solani* AG1--1a previously isolated from infected rice was maintained on potato dextrose agar (PDA) medium in tubes. The isolate was transferred to 90&nbsp;mm Petri dishes containing PDA and incubated a room temperature (20 to 27&deg;C). Glass bottles of autoclaved rice grain and hull substrate were prepared and plugs of the culture were transferred from Petri dishes to the autoclaved substrate and incubated at room temperature for two weeks.

The plots were inoculated 20 days after transplanting by spreading 151&nbsp;g of inoculum over the split--plot area with a 1&nbsp;m buffer on either end (4&nbsp;m x 11&nbsp;m).

@@

The 2016 experiment followed the same split--plot design as described for the 2015 experiment. However, the plot size was increased and due to these changes, the replicate sizes are are not equal as necessitated by the use of a larger area for the experiment. The main plot sizes were: Block 1 (B1) 21&nbsp;m x 20.5&nbsp;m (412.5 sq&nbsp;m) and Block 2 (B2) 20.25&nbsp;m x 21.6&nbsp;m (437.4&nbsp;m^2^). The split--plot sizes were B1 21&nbsp;m x 10.25&nbsp;m (215.25&nbsp;m^2^), B2 20.25&nbsp;m x 10.8&nbsp;m (218.7&nbsp;m^2^). The replication sizes were B1 --- 42&nbsp;m&nbsp;x&nbsp;20.5&nbsp;m (861&nbsp;m^2^) and B2 --- 40.5&nbsp;m&nbsp;x&nbsp;21.6&nbsp;m (874.8&nbsp;m^2^). A buffer 0.5&nbsp;m per split--plot was used and the overall experiment size was 3471.6&nbsp;m^2^.

split--plot N rates differed from the 2015 rates with only two N treatments, N60 (60&nbsp;kg&nbsp;/&nbsp;ha as urea split into two applications) and N180 (180&nbsp;kg&nbsp;/&nbsp;ha in three splits) being applied (Tab.&nbsp;\@ref(tab:N_rate)).

@@

In both experiments, two sample areas, sized 1&nbsp;m&nbsp;x&nbsp;1&nbsp;m, per plot were assessed. The total number of tillers per hill and number of tillers with sheath blight (incidence) were measured for nine hills per sample area. Tiller sheath severity was measured for four tillers per hill and six leaves tiller using a disease assessment rating scale for field evaluation developed at IRRI. The scale is an unequally spaced categorical scale where 0 -- no disease; 1 -- trace to 1&nbsp;\% severity; 2 -- 1 to 5&nbsp;\% severity; 3 -- 5 to 15&nbsp;\% severity; 4 -- 15 to 50&nbsp;\% severity; 5 -- 50 to 100&nbsp;\% severity. Severity was considered to be the amount of leaf or sheath tissue covered by sheath blight lesions. Leaves were rated as green (living) having at least XXX&nbsp;\% living tissue or dead, defined as having less than XXX&nbsp;\% living tissue. The same rating methodology was used in both experiments. Five disease assessments were made in the 2015 experiment and four disease assessments were made in the 2016 experiment, respectively.

@@

As most of the data's residuals did not meet assumptions for normality, the analysis was carried out using Bayesian multivariate generalised linear mixed models implemented in the R package `MCMCglmm` [@MCMCglmm2010]. Six models were created where the the dependent variables were 2015 tiller sheath blight incidence, 2015 tiller sheath blight severity, 2015 leaf sheath blight severity; 2016 tiller sheath blight incidence, 2016 tiller sheath blight severity; 2016 leaf sheath blight severity, respectively. The independent variables were the irrigation management and N rate with replicate treated as a random variable. The base levels (control treatments) for the analyses for 2015 were: N rate -- N0, irrigation management -- puddled and for 2016 they were: N rate -- N60, irrigation management -- puddled

Four MCMC chains were run for 55,000 steps with the first 5000 discarded as burn in. The remaining 50,000 samples were used to determine the posterior distribution of the parameters of the model. The `coda` [@coda2006] package was used to provide utilities to check MCMC chain convergence using visual inspection of resulting trace graphs and the Gelman--Rubin test.

@@

The combination of the irrigation and N rates did not appear to increase the disease severity or incidence in either experiment. In 2015 there were no clear differences in the interaction of irrigation and N rate between any of the three treatments. However, in the 2016 experiment the combination of AWD--N180 was clearly different, being lower, than the PDL--N60 combination treatment.

@@

Climate change is predicted to decrease sheath blight risk in the Korean peninsula over the longer term. However, over the next 10 -- 20 years the risk is predicted to remain the same as from 2000 -- 2010 [@kim2015predicting]. Therefore, it is likely to remain an important diease in the near future with little resistance available in current varieties. However, based on these studies the use of AWD technologies would not appear to be a factor that increases the incidence or severity of rice sheath blight disease.

@@

```{r fig-3, fig.cap="Sheath blight progress was measured in the 2015 at five assessment points and 2016 experiment at four points. Sheath blight incidence was rated as the number of infected tillers divided by the total number of tillers per hill and used to calculate the area under the disease progress stairs (AUDPS). Sheath blight severity was rated on an unevenly spaced categorical scale and converted to the mid--point percentage value to calculate the AUDPS. Main plot irrigation treatments were permanently puddled (PDL) and alternate wetting and drying (AWD)."}

@@

##### Tiller sheath blight incidence boxplot for N rate
```{r fig-4, fig.cap="Sheath blight severity progress was measured in the 2015 at five assessment points and 2016 experiment at four points. Sheath blight incidence was rated as the number of infected tillers divided by the total number of tillers per hill and used to calculate the area under the disease progress stairs (AUDPS). Sheath blight severity was rated on an unevenly spaced categorical scale and converted to the mid--point percentage value to calculate the AUDPS. Five different N treatments were applied to split plots depending on the experiment. In the 2015 experiment three N rate treatments were applied: no N (N0), 100&nbsp;kgkg&nbsp;/^&nbsp;ha N (N100) and 180&nbsp;kgkg&nbsp;/^&nbsp;ha N (N120). In the 2016 experiment, two N rate treatments were applied: 60&nbsp;kgkg&nbsp;/^&nbsp;ha N (N60) and 180&nbsp;kgkg&nbsp;/^&nbsp;ha N (N180)."}
